\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-14 k-planar.tex
added subsection on framed graphs

We first formally define $k$-planar graphs.  An \defin{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.  The \emph{interior} of $f$ is the point set $\{f(t):0<t<1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \defin{crossing} in an embedded (multi)graph $G$ is an unordered pair of distinct edges $\{vw,xy\}\subseteq E(G)$ whose interiors have a point in common. Without any loss of generality we may assume that the interiors of any two edges of $G$ have at most one point in common and that no point in $\R^2$ is contained in the interior of three or more edges of $G$.  This assumption can be enforced by local changes that do not change the number of crossings each edge is involved in.  If $p$ is a common point in the interior of two crossing edges $vw$ and $xy$ then we say that $vw$ and $xy$ \defin{cross at} $p$.

\subsection{\boldmath $d$-Framed and $1$-Planar Graphs}
\label{sec-1-planar}

In the important special case of 1-planar graphs we obtain better constants and an additional property (planarity) of $H$.

% \note{PM}{Include in note to referee:  We choose to keep this section because
%   $1$-planar graphs are of special interest to the graph drawing community.  For example, the annotated bibliography on $1$-planarity (\htmladdnormallink{https://arxiv.org/abs/1703.02261}{arXiv:1703.02261}) has 143 entries, and even this is now four years out of date.
%
%   An important aspect of our result for $1$-planar graphs is that graph $H$ in the product is planar and has treewidth $3$, i.e., it has simple treewidth $3$.  In a more recent application of product structure \citet{BDJM} show that such graphs have $\ell$-vertex rankings using $O(\log n/\log\log\log n)$ colours, and this is tight. This result requires that $H$ have simple treewidth $3$, even treewidth $3$ is not sufficient.
% }
%
% \note{DW}{I will rewrite the intro to this subsection once we decide what our results are.}

We will actually prove this result in the more general setting of $d$-framed graphs \cite{BDGGMR}. For any integer $d\ge 3$, a (multi)graph $G$ is \defin{$d$-framed} if it admits a drawing having a biconnected plane skeleton (multi)graph $G_0$, whose faces have size at most $d$, and whose crossing edges are in the interior of these faces.  In \cref{1_planar_is_4_framed}, below, we show that every $1$-planar graph is a subgraph of a $4$-framed multigraph. This makes \cref{1-planar} an immediate consequence of the following theorem, whose proof is the subject of the rest of this section.

\begin{thm}\label{d_framed_product_stucture}
  Every $d$-framed graph is contained in $H\boxtimes P\boxtimes K_{\lfloor d/2\rfloor + d - 3}$ for some planar graph $H$ with treewidth at most 3 and for some path $P$.
\end{thm}

\subsection{\boldmath $d$-Framed Graphs}

The following definition is a convenient way to estalish the planarity of $H$ in \cref{d_framed_product_stucture}:  A tree-decomposition $\mathcal{T}$ of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $t$-element subset $\{v_1,\ldots,v_t\}\subseteq V(G)$, at most two bags of $\mathcal{T}$ contain $\{v_1,\ldots,v_t\}$.  Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ be a $d$-framed graph and let $G'$ be the frame of $G$.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that $V(F)$ can be partitioned into $k\le 3$ sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
		\begin{compactenum}
			\item $F[P_i]$ is a path; and
			\item $P_i=Y_i\cup X_i$ where $|X_i|\le d-3$ and $|Y_i\cap L_j| \le 3$ for each $j\ge 0$.
		\end{compactenum}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in the closure of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
	\begin{compactenum}[(i)]
		\item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
        \item For each $x\in V(H)$, $S_x=Y_x\cup X_x$ where $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
		\item $H$ has a $3$-simple tree decomposition $\mathcal{T}$ and, if $k=3$, then $\{x_1,x_2,x_3\}$ is contained in exactly one bag of $\mathcal{T}$.
	\end{compactenum}
\end{lem}

@@

\subsection{\boldmath $1$-Planar Graphs}

In the following, we assume that no two edges incident to a common vertex of the $1$-plane graph $G$ cross each other since, in a 1-plane graph, such a crossing can always be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross.\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}  The \defin{faces} of $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$, where we treat each edge $vw$ of $G$ as the image of a simple curve $f_{vw}:[0,1]\to\R^2$.
% We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.

@@

\note{DW}{I am pretty sure one can get treewidth 3 for $(g,1)$-planar graphs, combining \cref{framed} with the treewidth 3 result in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). }

@@

\note{DW}{I am pretty sure one can get treewidth 3 here using \cref{framed} and the treewidth 3 result in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}).}

@@

@@

\begin{lem}
\label{NewLemma}
Let $G_0$ be a 2-connected plane multigraph. Let $T$ be a spanning tree of $G_0$ rooted at some vertex $r$ on the boundary of the outerface of $G_0$. Let $C = [P_1,\dots,P_k]$ be a cycle in $G_0$ with $r$ in the exterior of $C$, where $k\in\{1,2,3\}$ and $P_1, \dots, P_k$ are bipods in $G_0$. Let $G:=G_0\blah{C}$. Then $G$ has a partition $\PP$ into tripods with respect to $T$ such that $P_1,\dots,P_k\in\PP$, and $G / \PP$ has a tree-decomposition $\mathcal{T}$ of width at most 3 such that for every facial cycle $F$ of $G$ there is a bag $B$ of $\mathcal{T}$ such that every part of $\PP$ that intersects $F$ is in $B$.
\end{lem}

@@

The proof is by induction on the number of faces of $G$. If $G$ has just one internal face (which is bounded by $C$), then the desired partition is $\PP=\{P_1,\ldots,P_k\}$ and the desired tree-decomposition of $G/\PP$ consists of a single bag $\{P_1,\ldots,P_k\}$ indexed by a 1-node tree. Now assume that $G$ has at least two internal faces.

@@

Consider $j\in\{1,\dots,m\}$. Let $G_j$ be the subgraph of $G$ consisting of $Z_j$ and all the edges and vertices of $G$ in the interior of $Z_j$. The outerface of $G_j$ is $Z_j$. Observe that for some $i\in\{1,2,3\}$, we have
$Z_j=[R_{i,j}^+,R_{i+1,j}^-,S_{i,j}]$, where
$R_{i,j}^+$ is some subpath of $R_i^+$,
$R_{i+1,j}^-$ is some subpath of $R_{i+1}^-$, and
$L_{i,j}$ is the path $Z_j-V(C)$ (which is contained in $Y$). \note{DW}{Add a figure of a good case and a weird case.}
Since $R_i^+$, $R_{i+1}^-$ and $L_i$ are bipods,
$R_{i,j}^+$, $R_{i+1,j}^-$ and $L_{i,j}$ are bipods.
By construction, $v_{i-1}$ is not in $G_j$ (where $v_0=v_3$).
Thus $V(G_j)\setminus V(Z_j) \subsetneq V(G)\setminus V(C)$, and we may apply induction to $G_j$.  Thus $G_j$ has a partition $\PP_j$ into tripods with respect to $T$ such that $R_{i,j}^+, R_{i+1,j}^-, L_{i,j}\in\PP_j$, and 	$G_j / \PP_j$ has a tree-decomposition  $\mathcal{T}_j=\{B_x:x\in V(D_j)\}$ of width at most 3 such that for every facial cycle $F'$ of $G_j$ there is a bag $B$ of $\mathcal{T}_j$ such that every part of $\PP_j$ that intersects $F'$ is in $B$. Since $C$ is a facial cycle of $G_j$, there is a bag $B_{u_j}$ of $G_j$ such that $P_1,\dots,P_k\in B_{u_j}$.

@@

>>>>

%%%%%%%%%%%%%%%%%%%
\subsection{Framed Graphs}

For any integer $d\ge 3$, a (multi)graph $G$ is \defin{$d$-framed} if it admits a drawing in the plane having a biconnected plane skeleton (multi)graph $G_0$, whose faces have size at most $d$, and whose crossing edges are in the interior of these faces.  This definition (for simple graphs) was introduced by \citet{BDGGMR}. Every $d$-framed graph can be described by a $(2,\binom{d}{2})$-shortcut system applied to a planar graph. Thus \cref{ShortcutProduct} is applicable. However, for framed graphs we prove the following stronger result, where the treewidth bound on $H$ is best possible.

\begin{thm}\label{d_framed_product_stucture}
Every $d$-framed graph is contained in $H\boxtimes P\boxtimes K_{d+3\floor{d/2}-3}$ for some planar graph $H$ with treewidth at most 3 and for some path $P$.
\end{thm}

This result is used below to obtain product structure theorems for 1-planar graphs and for map graphs. For 1-planar graphs it is essential that we allow skeletons with parallel edges.

\cref{d_framed_product_stucture} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}. The following definition is a convenient way to establish the planarity of $H$ in \cref{d_framed_product_stucture}:  A tree-decomposition $\mathcal{T}$ of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $t$-element subset $\{v_1,\ldots,v_t\}\subseteq V(G)$, at most two bags of $\mathcal{T}$ contain $\{v_1,\ldots,v_t\}$.

\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ be a $d$-framed graph and let $G'$ be the frame of $G$.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that $V(F)$ can be partitioned into $k\le 3$ sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
		\begin{compactenum}
			\item $F[P_i]$ is a path; and
			\item $P_i=Y_i\cup X_i$ where $|X_i|\le d-3$ and $|Y_i\cap L_j| \le 3$ for each $j\ge 0$.
		\end{compactenum}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in the closure of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
	\begin{compactenum}[(i)]
		\item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
		\item For each $x\in V(H)$, $S_x=Y_x\cup X_x$ where $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
		\item $H$ has a $3$-simple tree decomposition $\mathcal{T}$ and, if $k=3$, then $\{x_1,x_2,x_3\}$ is contained in exactly one bag of $\mathcal{T}$.
	\end{compactenum}
\end{lem}

@@

We first formally define $k$-planar graphs.  An \defin{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.  The \defin{interior} of $f$ is the point set $\{f(t):0<t<1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior. The \defin{faces} of $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$. A \defin{crossing} in an embedded (multi)graph $G$ is an unordered pair of distinct edges $\{vw,xy\}\subseteq E(G)$ whose interiors have a point in common. Without any loss of generality we may assume that the interiors of any two edges of $G$ have at most one point in common and that no point in $\R^2$ is contained in the interior of three or more edges of $G$.  This assumption can be enforced by local changes that do not change the number of crossings each edge is involved in.  If $p$ is a common point in the interior of two crossing edges $vw$ and $xy$ then we say that $vw$ and $xy$ \defin{cross at} $p$.

\subsection{\boldmath $1$-Planar Graphs}
\label{sec-1-planar}

$1$-planar graphs are of particular interest to the graph drawing community~\citep{kobourov.liotta.ea:annotated}. In this case, we obtain better constants and an additional property (planarity) of $H$.

%(\htmladdnormallink{https://arxiv.org/abs/1703.02261}{arXiv:1703.02261}) has 143 entries, and even this is now four years out of date.
%
%   An important aspect of our result for $1$-planar graphs is that graph $H$ in the product is planar and has treewidth $3$, i.e., it has simple treewidth $3$.  In a more recent application of product structure \citet{BDJM} show that such graphs have $\ell$-vertex rankings using $O(\log n/\log\log\log n)$ colours, and this is tight. This result requires that $H$ have simple treewidth $3$, even treewidth $3$ is not sufficient.
% }

%In \cref{1_planar_is_4_framed}, below, we show that every $1$-planar graph is a subgraph of a $4$-framed multigraph. This makes \cref{1-planar} an immediate consequence of the following theorem, whose proof is the subject of the rest of this section.

\cref{1-planar} is an immediate consequence of \cref{d_framed_product_stucture} and the next lemma, which is of independent interest. We assume that no two edges incident to a common vertex of the $1$-plane graph $G$ cross each other since, in a 1-plane graph, such a crossing can always be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross.\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}  % We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.

\note{DW}{Say something about how \cref{1_planar_is_4_framed} is better than similar results in the literature.}

@@

\note{DW}{I am pretty sure one can get treewidth 3 for $(g,1)$-planar graphs, combining \cref{framed} with the treewidth 3 result in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}

@@

\note{DW}{I am pretty sure one can get treewidth 3 here using \cref{framed} and the treewidth 3 result in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}

@@

\note{DW}{Work in progress}

@@

\begin{lem}
\label{NewLemma}
Let $G_0$ be a 2-connected plane multigraph. Let $T$ be a spanning tree of $G_0$ rooted at some vertex $r$ on the boundary of the outerface of $G_0$. Let $C = [P_1,\dots,P_k]$ be a cycle in $G_0$ with $r$ in the exterior of $C$, where $k\in\{1,2,3\}$ and $P_1, \dots, P_k$ are bipods in $G_0$. Let $G:=G_0\blah{C}$. Then $G$ has a partition $\PP$ into tripods with respect to $T$ such that:
\begin{compactenum}[(i)]
	\item $P_1,\dots,P_k\in\PP$, 
	\item $G / \PP$ has a tree-decomposition $\mathcal{T}$ of width at most 3,
	\item $G/\PP=\widehat{G}/\PP$.
%	 such that for every facial cycle $F$ of $G$ there is a bag $B$ of $\mathcal{T}$ such that every part of $\PP$ that intersects $F$ is in $B$.
\end{compactenum}
\end{lem}

@@

The proof is by induction on the number of faces of $G$. Suppose $G$ has just one internal face (which is bounded by $C$). Let $\PP=\{P_1,\ldots,P_k\}$.
Then $G/\PP=\widehat{G}/\PP = K_k$. Then desired tree-decomposition of $G/\PP$ consists of a single bag $\{P_1,\ldots,P_k\}$ indexed by a 1-node tree. Now assume that $G$ has at least two internal faces.

@@

Consider $j\in\{1,\dots,m\}$. Let $G_j$ be the subgraph of $G$ consisting of $Z_j$ and all the edges and vertices of $G$ in the interior of $Z_j$. The outerface of $G_j$ is $Z_j$. Observe that for some $i\in\{1,2,3\}$, we have
$Z_j=[R_{i,j}^+,R_{i+1,j}^-,S_{i,j}]$, where
$R_{i,j}^+$ is some subpath of $R_i^+$,
$R_{i+1,j}^-$ is some subpath of $R_{i+1}^-$, and
$L_{i,j}$ is the path $Z_j-V(C)$ (which is contained in $Y$). \note{DW}{Add a figure of a good case and a weird case.}
Since $R_i^+$, $R_{i+1}^-$ and $L_i$ are bipods,
$R_{i,j}^+$, $R_{i+1,j}^-$ and $L_{i,j}$ are bipods.
By construction, $v_{i-1}$ is not in $G_j$ (where $v_0=v_3$).
Thus $V(G_j)\setminus V(Z_j) \subsetneq V(G)\setminus V(C)$, and we may apply induction to $G_j$.  Thus $G_j$ has a partition $\PP_j$ into tripods with respect to $T$ such that 
\begin{compactenum}[(i)]
	\item
	$R_{i,j}^+, R_{i+1,j}^-, L_{i,j}\in\PP_j$, 
	\item $G_j / \PP_j$ has a tree-decomposition $\mathcal{T}_j=\{B_x:x\in V(D_j)\}$ of width at most 3,
	\item $G_j / \PP=\widehat{G_j}/\PP$.
	%	 such that for every facial cycle $F$ of $G$ there is a bag $B$ of $\mathcal{T}$ such that every part of $\PP$ that intersects $F$ is in $B$.
\end{compactenum}

@@

%Since $P_1,
%Since $C$ is a facial cycle of $G_j$, there is a bag $B_{u_j}$ of $G_j$ such that $P_1,\dots,P_k\in B_{u_j}$.
